---
title: "Exempelblad dör datalab"
author: "Anders Johansson"
date: '2020-09-30'
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(flextable)
library(fosdata)

knitr::opts_chunk$set(echo = TRUE,warning= FALSE)
theme_set(theme_light())
set_flextable_defaults(digits=4,theme_fun = theme_zebra)
use_df_printer()
```


Följande arbetsblad innehåller lite blandade och oredigerade anteckningar för uppgifterna i datorlabben. 


# T-test för skillnad mellan två medel

8.29 Cigaretter

The data set cigs from the fosdata package contains measurements of cigarette brands that were tested by the U.S. Federal Trade Commission for chemical content. Check with a plot that the nicotine (nic) and tar variables are reasonably bump shaped and symmetric, without major outliers.

- Test for a difference in nicotine content between filtered and unfiltered cigarette brands.  State your result with a p-value.
- Test for a difference in tar content between filtered and unfiltered cigarette brands. State  your result with a p-value.
- Test for a difference in nicotine content between menthol and non-menthol cigarette brands.  State your result with a p-value.
- Test for a difference in tar content between menthol and non-menthol cigarette brands. State  your result with a p-value.



```{r}
cigs <- fosdata::cigs   # Remove missing values
cigs %>% sample_n(10)
```

Boxplot av nikotinhalt filter/icke-filter

```{r}
cigs %>% 
  ggplot(aes(x=filter,y=nic,fill=filter)) +
  geom_boxplot()
```
Tjärhalt (`tar`)
```{r}
cigs %>% 
  ggplot(aes(x=filter,y=tar,fill=filter)) +
  geom_boxplot()
```
```{r}
cigs %>% ggplot(aes(sample=nic,color=filter)) +
  geom_qq()
```

```{r}
t.test(nic ~ filter,data=cigs,conf.level=0.95)
```

Kolla normalfördelningsantagandet
```{r}
cigs %>% ggplot(aes(sample=tar,color=filter)) +
  geom_qq()
```
Kommentar: Ej troligt normalfördelningsantagandet gäller. 

```{r}
t.test(tar ~ filter,data=cigs,conf.level=0.95)
```

## 8.28 Benstyrka 
Detta är ett exempel på “stickprov i par” vilket egentligen betyder att man behandlar det som ett stickprov där differenserna.

Experimenters measured the leg strength of adults using two methods: a stationary isometric dynamometer and a Wii Balance Board. The strength (in kg) of each adult is reported in fosdata::leg_strength.

Test for a difference in the mean leg strength as measured with these two devices.

```{r}
leg_strength <- fosdata::leg_strength

leg_strength %>% sample_n(10)
```
```{r}
leg2 <- leg_strength %>% mutate(differens=mean_wii-mean_sid) 
leg2
```

```{r}
leg2 %>% ggplot(aes(sample=differens)) +
  geom_qq() +
  geom_qq_line()
```

```{r}
t.test(leg2$differens)
```


# Chi-två test

# 10.17 
Consider the =fosdata::normtemp= data set.

Use a goodness of fit test with 10 bins, all with equal probabilities,
to test the normality of the temperature data set. Note that in this
case, you will need to estimate two parameters, so the degrees of
freedom will need to be adjusted appropriately.

Läs in data i `df1`.

```{r}
df1 <- mutate(fosdata::normtemp, gender=ifelse(gender == 1, "M","F"))
df1 %>% sample_n(20)
nrow(df1)
```


Vi försöker undersöka en normalmodell $N(\mu,\sigma)$ : Skattar $\mu$
och $\sigma$ från data.

```{r}
mu1 <- mean(df1$temp) # skatta 
sd1 <- sd(df1$temp)
```

```{r}
p <- ggplot(df1,aes(x=temp)) +
  geom_histogram(aes(y=..density..),bins=15,fill="steelblue",alpha=0.5) +
  geom_function(fun=function(x) dnorm(x,mean=mu1,sd=sd1))
p
```


```{r}
breaks <- c(0,qnorm(seq(0.1,0.9,by=0.1),mean=mu1,sd = sd1),1000)
breaks
```


```{r}
df1 %>% 
  mutate(category=cut(temp,breaks),count=1) %>% 
  group_by(category) %>%
  summarise(O=sum(count),E=0.1*nrow(df1)) %>% ungroup() -> counts
counts 
```




Vi måste justera p-värdet eftersom vi skall använda $\nu=10-1-2$
frhetsgrader. (2 Skattade parametrar.) Vi beräknar här nere $\chi^2$
värdet direkt.

```{r}
chi <- summarise(counts,X = sum( (O-E)^2/E ),pvalue=1-pchisq(X,df=10-1-2))
chi
```

Vi förkastar **inte** nollhypotesen eftersom p-värdet är så högt (69%).

# Enkel och multipel linjär regression

Läs översiktligt 
[kapitel 11](https://mathstat.slu.edu/~speegle/_book/SimpleReg.html#SimpleReg) 
i
PSD om enkel linjär regression fram till 11.7. Gör en av uppgifterna
11.18 och 11.39. Läs översiktligt igenom kapitel 13 och gör en av
uppgifterna 13.4 och 13.8

I rapporterna för regressionsuppgifterna bör ni undersöka
fördelningsantaganden genom att studera residualerna. 

Genomför en kort så kallad [residualanalys](https://mathstat.slu.edu/~speegled/_book/SimpleReg.html#residualanalysis)
som i avsnitt 11.4: Ni kan använda qq-plot (`geom_qq()`
eller helt enkelt anropa `plot(m)` där `m`
returneras av `lm` eller `glm`. Kommentera kort
utseendet på residualernas fördelning.

## 11.18 Starwarsfigurer

Consider the `starwars` data set in the `dplyr`
package.

1.  Create a scatterplot of mass versus height of the *Star Wars*
    characters.
2.  Create a linear model of mass as described by height for the *Star
    Wars* characters, and write the equation for the line of best fit.
3.  Examine the residual plots and comment. Which character is the
    outlier?
4.  Remove large outliers and create a new linear model of mass as
    described by height for the *Star Wars* characters minus the
    outlier. Comment on the magnitude of the change of the slope and
    intercept.
    
Presentera data
```{r}
starwars %>% filter(!is.na(mass)) -> sw
```

En scatterplot av vikt mot längd.
```{r}
sw %>% 
  ggplot(aes(x=height,y=mass)) +
  geom_point() +
  geom_smooth(formula=y~x,method="lm")
```
Använder `lm`för att beräkna regressionsmodellen

```{r}
lm_starwars <- lm(mass~height,data=sw)
summary(lm_starwars)
```
```{r}
plot(lm_starwars,1)
```
```{r}
plot(lm_starwars,2)
```

Jabba är outlier

```{r}
sw[15,]
```

```{r}
sw2 <- filter(sw,mass < 1000)
sw2 %>% 
  ggplot(aes(x=height,y=mass)) +
  geom_point() +
  geom_smooth(formula=y~x,method="lm")
```


```{r}
lm2 <- lm(mass~height,data=sw2)
summary(lm2)
```

```{r}
plot(lm2,1)
```
```{r}
plot(lm2,2)
```



11.39 Icke-linjär modell

I den här uppgiften kan ni hoppa över beräkningarna av \"MSE LOOCV\" som
ingår.

Example 11.4 used a regression line to model speed on the \"shape trail
B\" test on age of children, with data from
`fosdata::child_tasks`. Rather than a linear relationship, an
inverse relationship of the form
$$ \hat{y} = \beta_{0} + \frac{\beta_{1}}{x} $$ may be a better model
for this data. You can build the inverse model using the formula
`stt_cv_trail_b_secs ~ I(1/age_in_months)`.

Make a plot showing the data, the linear model, and the inverse model.
(Hint: use `predict` on the inverse model to create predicted
$y$-values for every data point and then plot them with
`geom_line`.)

13.5

The built-in data set `swiss` contains a standardized
fertility measure and socio-economic indicators for each of 47
French-speaking provinces of Switzerland at about 1888.

```{r}
swiss
```


- Investigate the distribution of the `Catholic` variable. What
do you observe? What does this tell you about Catholics and Protestants
in 19th century Switzerland?

```{r}
swiss %>% ggplot(aes(x=Catholic)) +
  geom_density()
```



- Produce a plot of `Fertility` as a function of
`Education`. Use the `cut` function to divide the
`Catholic` variable into three classes, and color your points
with three colors for those three classes.

```{r}
swiss %>% mutate(Catholic=cut(Catholic,breaks=c(0,25,75,100))) %>% 
  ggplot(aes(x=Education,y=Fertility,color=Catholic)) +
  geom_point()
```


- Form a linear model of `Fertility`? on the other five
variables. Which variables are significant at the 0.05 level?

```{r}
swissmod <- lm(Fertility~Agriculture+Examination+Education+Catholic+Infant.Mortality,data=swiss)
summary(swissmod)
```


- Drop any variables that are not significant and make a new linear model.
  How does this change the adjusted $R^2$
```{r}
swissmod <- lm(Fertility~Agriculture+Education+Catholic+Infant.Mortality,data=swiss)
summary(swissmod)
```



13.8

The `penguins` data from the `palmerpenguins`
package has body measurements of three species of penguin.

Make a scatterplot of `body_mass_g` as a function of
`flipper_length_mm`, color by `species`, and show
the regression lines from the variable slopes model (just use
`geom_smooth` with `method=lm`).

Fit a linear model of `body_mass_g` on
`flipper_length_mm`, `species`, and include the
interaction between `flipper_length_mm` and
`species`.

Which interaction terms are significant? Explain your answer based on
the plot.



